---
title: "data_01_exploration"
author: "Jennifer Doyle"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    number_sections: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r libs, message=FALSE, echo=TRUE, warning=FALSE}
gc()
rm(list=ls())

library(RODBC)
library(tidyverse)
library(lattice)
library(lubridate)
library(mapplots)
library(sp)
library(sf)
library(reshape2)
library(readxl)
library(knitr)
library(DataCombine)
library(lattice)
library(icesTAF)
library(magrittr)
library(viridis)
library(sessioninfo)
library(stringr)
```
# Setup options

```{r setupoptions}
curr.year <- 2025
dat.year <- curr.year -1
fu.n <- "FU19"

save.plots <- T # set it up as T or F whether you want to save the plots when running all the chunks or kniting the document
save.tables <- T # set it up as T or F whether you want to save the tables "propmature"" and the "L50" into .txt files when running all the chunks or knitting the document

# ewen plot
source("boot/software/functions/required.funcs.r")
source("boot/software/functions/ld1.plot.r")


save.plots <- T # set it up as T or F whether you want to save the plots when running all the chunks or kniting the document
save.tables <- T # set it up as T or F whether you want to save the tables "propmature"" and the "L50" into .txt files when running all the chunks or knitting the document
```

# Introduction

The goals of this R Markdown document are to:

* Preprocess and explore FU19 Intercatch data
* Write TAF data tables

Files before (inside "boot/data"):

* intercatch folder
  * fu19_int.land.csv 
  * fu19.nat.qlfd.all.wg2024.csv
  * MSY_nep_stocks.csv
  * NepLand_WGCSE2024.csv
  * PropRet_WKCELT_2014.csv
  * StockOverview_FU19.txt


Files after (inside "data/data_01_exploration"):

* nat.qlfd.wg_FU19.csv
* discard.data_FU19.csv
* discard.data.sex_FU19.csv
* prop.male.wgcse_FU19.csv
* national_landings_FU19.csv
* international.landings_FU19.csv


* fu19_WKCELT_discardogive.png
* fu19_Discard Rate.national.programme.png
* fu19_Discard Rate.national.programme.sex.png
* fU19_ProportionMale.png
* fu19_Proportion Male_quarterly.png
* fu19_NationalLandings.png
* fu19_International Landings.png
* fu19_LFD_plot.png

# Data Work Up
<a href="#top">Back to top</a>

This Rmarkdown document gives the exploitation pattern from the raised numbers from Irish Sampling programme.
Where this aggregated data is investigated for WG stock summary

Outputs various WG report figures and tables.

## Load data from 01_FU19_Extraction_WGCSE_2024
<a href="#top">Back to top</a>

Load in national raised QLFD object for 2003-2024. 

Load in data year raised QLFD object for Intercatch data submission.
All raised quarterly using annual discard ogive derived at WKCELT 2014 and follows Stock Annex.
Reference document ""01_FU19_Extraction_IntercatchFormat_WGCSE2023.rmd" on national network.

```{r data load, echo=FALSE, results='true'}

qlfd.wg  <- read.csv("boot/data/intercatch/fu19.nat.qlfd.all.taf.wg2025.csv")
qlfd.wg<-  qlfd.wg[ , c(-1)]
#qlfd.hist <-  qlfd.hist %>% select (year, q, cl, sex, rdis.n.r, rlan.n.r, rdis.wt.r, rlan.wt.r)


fileOut <- paste("data/data_01_exploration/nat.qlfd.wg_", fu.n, ".csv", sep = "")
 if (save.tables == T) {
   write.table(qlfd.wg, fileOut, sep=",", row.names = F)
 }
```

## Derived Annual Ogive WKCELT 2014
<a href="#top">Back to top</a>

```{r dig ogive, echo=TRUE, fig.height=7, fig.width=7, warning=FALSE, fig.cap="WKCELT derived discard ogive."}


df.hist  <- read.csv("boot/data/intercatch/PropRet_WKCELT_2014.csv")

#rename current wg year
df.hist$Year <- "2023"
df.hist$Year <- as.numeric(df.hist$Year)


names(df.hist) %<>% tolower

pr <- ggplot() +
      geom_line(data=df.hist, aes(x=clmm, y= propret), colour = 'blue', size = 1, linetype = "twodash") +
      theme_bw() + ylab("Proportion Retained") + ggtitle("WKCELT Ogive")

pr + geom_hline(yintercept=0.5) +
  theme(strip.text.x = element_text(size=14, face="bold"))

if (save.plots == T) {
  ggsave("data/data_01_exploration/fu19_WKCELT_discardogive.png")
}
```


## Plot the estimated discard rates by number and weight from the national sampling programme.
<a href="#top">Back to top</a>

```{r dis data1, echo=TRUE, fig.height=7, fig.width=7, warning=FALSE, fig.cap="Discard rates by  number (dark-line) and weight (light-line) from national sampling programme."}

viridis_colors <- viridis_pal()(3)

dis <- qlfd.wg %>% group_by(year) %>% summarise( lan.wgt=sum(rlan.wt.r),
                                                 dis.wgt=sum(rdis.wt.r),
                                                lan.num=sum(rlan.n.r),
                                                dis.num=sum(rdis.n.r)) %>% 
                                      mutate(dis.rate.wt=dis.wgt/(lan.wgt+dis.wgt),
                                             dis.rate.n=dis.num/(lan.num+dis.num),
                                             lan.mw=lan.wgt/lan.num*1000,
                                             dis.mw=dis.wgt/dis.num*1000)
knitr::kable(dis , digits=3)

fileOut <- paste("data/data_01_exploration/discard.data_", fu.n, ".csv", sep = "")
 if (save.tables == T) {
   write.table(dis, fileOut, sep=",", row.names = F)
 }


#gather data and plot

dis1 <- dis[c(1, 6:7)]
dis1 <- dis1 %>% gather("disw","var" ,2:3)
names(dis1) <- c("year", "type", "rate")

p1 <- ggplot(dis1, aes(year, rate, group = type, colour=factor(type))) + 
  geom_line(size = 1) +
  ylab("% Discard rate") +
  theme_bw() +
  coord_cartesian(ylim= c(0, 1), xlim=c(2008,dat.year)) +
  scale_x_continuous(name="\nYear",
                     breaks = seq(min(as.numeric(as.character(dis1$year))),
                                  max(as.numeric(as.character(dis1$year))), by = 1)) +
  scale_colour_viridis_d(name = "Type", option = "D", end = 0.9) +  # Using color-blind friendly palette
  theme(axis.text=element_text(size=11),
        axis.title=element_text(size=11),
        legend.title=element_blank(),
        panel.grid=element_blank(),
        legend.position = "bottom")
p1

if (save.plots == T) {
  ggsave("data/data_01_exploration/fu19_Discard Rate.national.programme.png")
}
```


## Now by sex
<a href="#top">Back to top</a>

```{r dis data2, echo=TRUE,fig.height= 7, fig.width=7, warning=FALSE, fig.cap="Discard rates by sex in number (dark-line) and weight (light-line) from national sampling programme."}

dis.sex <- qlfd.wg %>% group_by(year, sex) %>% summarise( lan.wgt=sum(rlan.wt.r),
                                                 dis.wgt=sum(rdis.wt.r),
                                                lan.num=sum(rlan.n.r),
                                                dis.num=sum(rdis.n.r)) %>% 
                                      mutate(dis.rate.byweight=dis.wgt/(lan.wgt+dis.wgt),
                                             dis.rate.bynumber=dis.num/(lan.num+dis.num))
knitr::kable(dis.sex , digits=3)

fileOut <- paste("data/data_01_exploration/discard.data.sex_", fu.n, ".csv", sep = "")
 if (save.tables == T) {
   write.table(dis.sex, fileOut, sep=",", row.names = F)
 }

#gather data and plot
dis.sex <- dis.sex[c(1:2, 7:8)]
dis.sex <- dis.sex %>% gather("disw","var" ,3:4)
names(dis.sex) <- c("year", "sex", "type", "rate")




p2 <- ggplot(dis.sex, aes(year, rate, group = type, colour=factor(type))) + 
      geom_line(size = 1) + facet_wrap(~sex) +
      ylab("Discard rate %") + theme_bw() +
      coord_cartesian(ylim= c(0, 1), xlim=c(2008,dat.year)) +
      theme(panel.grid=element_blank(), legend.position = "bottom") +
      scale_x_continuous(name="\nYear",
                     breaks = seq(min(as.numeric(as.character(dis.sex$year))),
                                 max(as.numeric(as.character(dis.sex$year))), by = 1)) +
    scale_colour_viridis_d(name = "Type", option = "D", end = 0.9) +  # Using color-blind friendly palette
    theme(axis.text=element_text(size=11),
        axis.title=element_text(size=11),
        legend.title=element_blank(),
        panel.grid=element_blank(),
        legend.position = "bottom")

p2

if (save.plots == T) {
  ggsave("data/data_01_exploration/fu19_Discard Rate.national.programme.sex.png")
}

```


## Plot the sex ratio to see if any trends as the fishery is male biased.
<a href="#top">Back to top</a>

```{r sr, echo=TRUE, fig.height=7, fig.width=7, warning=FALSE, fig.cap="% of Male in Catches and Landings from national sampling programme."}

pd <- qlfd.wg %>% group_by(year, sex) %>% summarise(lan.num=sum(rlan.n.r),
                                           dis.num=sum(rdis.n.r))  %>%
                                      mutate(cat.num= lan.num+dis.num)

pr <- pd %>% gather("type", "nb", 3:5)

#prop male in landings
tln <- pr %>% filter(type=="lan.num") %>% group_by(year) %>% summarise(n=sum(nb))
tlm <- pr %>% filter(type=="lan.num", sex =="Male") %>% group_by(year) %>% summarise(n=sum(nb))
prop.m.ln <-left_join(tlm,tln, by="year") %>% mutate(prop.male.land = (n.x/n.y))


#prop male in catches
tcn <- pr %>% filter(type=="cat.num") %>% group_by(year) %>% summarise(n=sum(nb))
tcm <- pr %>% filter(type=="cat.num", sex =="Male") %>% group_by(year) %>% summarise(n=sum(nb))
prop.m.cn <-left_join(tcm,tcn, by="year") %>% mutate(prop.male.catch = (n.x/n.y))
pro <-left_join(prop.m.cn,prop.m.ln, by="year") 
pro <- pro[c(1, 4,7)]

knitr::kable(pro, digits=2) 

fileOut <- paste("data/data_01_exploration/prop.male.wgcse_", fu.n, ".csv", sep = "")
 if (save.tables == T) {
   write.table(pro, fileOut, sep=",", row.names = F)
 }


p3 <- ggplot(data = pro, aes(x = year)) +
  geom_line(aes(y= prop.male.catch, colour = "Catch"), size = 1) + 
  geom_line(aes(y= prop.male.land, colour = "Landings"), size = 1) +
  theme_bw() + ylab("% male by number")  +
  scale_colour_manual("", 
                      values = c("Catch"="blue", "Landings"="black")) +
  coord_cartesian(ylim= c(0, 1), xlim=c(2008, dat.year)) +
  theme(panel.grid=element_blank(), legend.position = "bottom") +
  scale_x_continuous(name="\nYear",
                     breaks = seq(min(as.numeric(as.character(pro$year))),
                                 max(as.numeric(as.character(pro$year))), by = 1)) +
    scale_colour_viridis_d(name = "Type", option = "D", end = 0.9) +  # Using color-blind friendly palette
    theme(axis.text=element_text(size=11),
        axis.title=element_text(size=11),
        legend.title=element_blank(),
        panel.grid=element_blank(),
        legend.position = "bottom")

p3

if (save.plots == T) {
  ggsave("data/data_01_exploration/fu19_ProportionMale.png")
}
```



## Plot the sex ratio by quarter.
<a href="#top">Back to top</a>

```{r srq, echo=TRUE, fig.height=7, fig.width=7,warning=FALSE, fig.cap="% of Male in Catches and Landings from national sampling programme by qtr."} 

pd <- qlfd.wg %>% group_by(year, q, sex) %>% summarise(lan.num=sum(rlan.n.r),
                                           dis.num=sum(rdis.n.r))  %>%
                                      mutate(cat.num= lan.num+dis.num)

pr <- pd %>% gather("type", "nb", 4:6)


#prop male in landings
tln <- pr %>% filter(type=="lan.num") %>% group_by(year, q) %>% summarise(n=sum(nb))
tlm <- pr %>% filter(type=="lan.num", sex =="Male") %>% group_by(year, q) %>% summarise(n=sum(nb))

# make join field
tln$yq <- as.character(tln$year*100 + tln$q)
tlm$yq <- as.character(tlm$year*100 + tlm$q)

prop.m.ln <-left_join(tlm,tln, by="yq") %>% mutate(prop.male.land = (n.x/n.y))


#prop male in catches
tcn <- pr %>% filter(type=="cat.num") %>% group_by(year, q) %>% summarise(n=sum(nb))
tcm <- pr %>% filter(type=="cat.num", sex =="Male") %>% group_by(year, q) %>% summarise(n=sum(nb))

# make join field
tcn$yq <- as.character(tcn$year*100 + tcn$q)
tcm$yq <- as.character(tcm$year*100 + tcm$q)


prop.m.cn <-left_join(tcm,tcn, by="yq") %>% mutate(prop.male.catch = (n.x/n.y))
proq <-left_join(prop.m.cn,prop.m.ln, by="yq") 
proq <- proq[c(1, 2, 8, 15)]
names(proq) <- c("year", "q", "prop.male.catch", "prop.male.land")

ggplot(data = proq %>% filter(year >=2008), aes(x = q)) +
  geom_line(aes(y= prop.male.catch, colour = "Catch"), size = 1) + 
  geom_line(aes(y= prop.male.land, colour = "Landings"), size = 1) +
  geom_point(aes(y= prop.male.catch, colour = "Catch"), size = 2) +
  geom_point(aes(y= prop.male.land, colour = "Landings"), size = 2) +
  theme_bw() + ylab("% male by number") + facet_wrap(~ year, nrow=3) +
    scale_colour_viridis_d(name = "Type", option = "D", end = 0.9) +  # Using color-blind friendly palette
    theme(axis.text=element_text(size=11),
        axis.title=element_text(size=11),
        legend.title=element_blank(),
        panel.grid=element_blank(),
        legend.position = "bottom")

if (save.plots == T) {
  ggsave("data/data_01_exploration/fu19_Proportion Male_quarterly.png")
}
```



## National Landings.
<a href="#top">Back to top</a>

Using the revised landings from 2008 - 2017.
At WGCSE 2019 Irish landings were revised back to 2008.
Revisions were in the order of +/- 5%. 
In one year it was 10%.
Also including revisions for sales notes under 10s in 2018.
Generate fishery exploitation pattern according to annex.

```{r nat land, echo=TRUE, message=FALSE, fig.height=10, fig.width=7, warning=FALSE, fig.cap="FU19 Irish  Landings (t) by quarter."} 

nat <- read.csv("boot/data/intercatch/NepLand_WGNEPH2025.csv")

nat$LiveWt <-as.numeric(as.character(nat$LiveWt))
nat$LiveWtRaisedToDecs <-as.numeric(as.character(nat$LiveWtRaisedToDecs))
nat$LiveWt <- nat$LiveWt/1000
nat$Quarter <- as.factor(nat$Quarter)
nat <- nat %>% filter(FU==19)

ggplot(nat, aes(Year, LiveWt))+ geom_col(aes(fill = Quarter)) + 
       theme_bw() + ylab("Landings in Tonnes") +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank()) +
  scale_x_continuous(breaks = c(seq(min(nat$Year), max(nat$Year), by = 5), max(nat$Year))) +
  scale_y_continuous(breaks = seq(0, 3800, 200)) 


if (save.plots == T) {
  ggsave("data/data_01_exploration/fu19_NationalLandings.png")
}

nl <- nat %>% group_by(Year, Quarter) %>% summarise(total=sum(LiveWt)) %>% spread(Quarter, total)
nl$Total<- rowSums(nl[,2:5], na.rm=TRUE)

knitr::kable(nl , digits=3)

fileOut <- paste("data/data_01_exploration/national.landings_", fu.n, ".csv", sep = "")
 if (save.tables == T) {
   write.table(nl, fileOut, sep=",", row.names = F)
 }
```


## Check if missing Catch sampling in time-series.
<a href="#top">Back to top</a>

There is one year 2019 q3 and earlier in time series as sampling was developing.
This has been accounted for in the national raising and the data object qlfd accounts for this.
Here is a check of this.

```{r check , echo=TRUE, warning=FALSE, fig.height=10, fig.width=7}
test <- qlfd.wg %>% group_by(year) %>% summarise(lan.num =sum(rlan.n.r), 
                                               lan.wgt =sum(rlan.wt.r),
                                               dis.num =sum(rdis.n.r),
                                               dis.wgt =sum(rdis.wt.r))


#raise to missing sample quarters as international data is not by quarter
check <- qlfd.wg %>% group_by(year) %>% summarise(rlan.wt=sum(rlan.wt.r), rdis.wt=sum(rdis.wt.r))
 
land1 <- nat %>% rename_at("Year",~"year") %>% group_by(year) %>% summarise(tot =sum(LiveWt))
test <- left_join(check, land1, by="year")
test$rf1 <- with(test, tot/rlan.wt)

knitr::kable(test , digits=5)
```




## InterNational Landings 
<a href="#top">Back to top</a>

The most up to date landings from each participant country is included here from Intercatch.

```{r int land, echo=TRUE, fig.height=10, fig.width=7, message=FALSE, fig.cap="FU19 International Landings (t)."}

int.land <- read.csv("boot/data/intercatch/fu19_int.land.csv")
int.land <- gather(int.land,"country", "landings",3:6) 


i <- int.land %>% filter(Year >= '1999')
pi <- ggplot(i, aes(Year, landings, fill=country)) + geom_bar(stat="identity", position= "stack") + 
  theme_bw() + ylab("Landings in Tonnes") +
  scale_fill_manual(breaks = c("France", "Belgium", "Ireland", "UK"),
                     values=c("#56B4E9","#E69F00","#009E73", "#CC6666")) +
  theme(panel.grid.major=element_blank(), panel.grid.minor=element_blank()) +
  scale_x_continuous(breaks = c(seq(min(int.land$Year), max(int.land$Year), by = 5), max(int.land$Year))) +
  scale_y_continuous(breaks = seq(0, 3800, 200))

pi + theme(axis.text=element_text(size=10),axis.title=element_text(size=10)) 


if (save.plots == T) {
  ggsave("data/data_01_exploration/fu19_International Landings.png")
}


int.land <- int.land %>% mutate_all(funs(replace(., is.na(.), 0)))

int.land  <- spread(int.land, country, landings)
int.land$total<- rowSums(int.land[,3:6], na.rm=TRUE)

knitr::kable(int.land , digits=2)

fileOut <- paste("data/data_01_exploration/international.landings.", fu.n, ".csv", sep = "")
 if (save.tables == T) {
   write.table(int.land, fileOut, sep=",", row.names = F)
 }
```


## Commercial Length Frequency Distributions
<a href="#top">Back to top</a>

Here we write the data  in the format for the ld1.plot.
The vertical lines indicate the minimum conservation reference size (25 mm) and the 35 mm visual reference level.

```{r ep , echo=TRUE, warning=FALSE, fig.height=8, fig.width=8, fig.cap="FU19 Commercial LFD."}

fu19.ld <- qlfd.wg[c(1:6)]
fu19.ld <- fu19.ld %>% gather("rdis.n.r","n" ,5:6)
names(fu19.ld) <- c("year", "q", "cl", "sex", "type", "n")

fu19.ld <- fu19.ld %>% group_by(Year= year, Length= cl, Sex=sex, Type=type) %>% summarise(n=sum(n))
fu19.ld <- spread(fu19.ld, Type, n)
fu19.ld$Catch <- fu19.ld$rdis.n.r  + fu19.ld$rlan.n.r
names(fu19.ld) <- c("Year",  "Length", "Sex", "Discards", "Landings", "Catch")
## here we write the CSV file in the fromat for the ld1.plot


df <- as.data.frame(fu19.ld)
#summary(df)

#the names of the data must be
#Year, Sex, Length, Landings, Discards, Catch
#in this instance they already are so we don't have to worry about it.
#call the function, parameter order is
#1 data frame
#2 title for the plot
#3 minimum year to start from
#4 maximum year to run to
#5 Minimum landing size
#6 reference line (typically 35mm so that we can see the proportion of the stock which is considered #fully selected.


png("data/data_01_exploration/fu19_LFD.plot.png", height=1700, width=1200, res=200)

plot.ld(df, "FU19", 2003, 2024, 25, 35)

dev.off()

plot.ld(df, "FU19", 2003, 2024, 25, 35)

```


## Session
<a href="#top">Back to top</a>
```{r info , echo=TRUE, warning=FALSE, fig.height=8, fig.width=8}
session_info()

```
